Modelling state-switching behaviour in functional data with hidden Markov models

### 2020-09-01 Report/Sections/models.tex
Clean up report and delete old files

Hidden Markov models describe state-switching Markovian processes in discrete time and will be the core structure of the models we use for $Y$ and $Y^*$. %Traditional HMMs model process auto-correlation by assuming that the underlying behaviour of a processes follows a single Markov chain. However, when conditioned on the hidden state sequence observations are assumed to be independent. Therefore, classical HMMs do not hold when the observations exhibit other forms of significant correlation in time. Examples of this include continuous-time processes such as a Wiener process, systems which exhibit temporal dependencies acting at several different time scales \citep{Barajas:2017}, and periodic behaviour including the motion of an ideal spring.  
%\textit{blah blah blah -quick restatements (5-15 words to potentially many sentences) of the main issues you are going to fix made a quick try at it, work on smoothing things. You can use some of the text I removed from below}.
%To account for each of these common additional dependence structures,
We first detail the structure of a traditional HMM. We then review a variation described by \cite{Lawler:2019} called the conditionally auto-regressive hidden Markov model (CarHMM), which explicitly models auto-correlation between observations. Either the HMM or the CarHMM can be used to model a sequence of observations at one of the two scales: for the coarse scale sequence $Y$ or for each of the fine scale sequences, $Y_1^*,\ldots, Y_T^*$. To simultaneously model $Y$ and $Y^*$, we use the hierarchical hidden Markov model (HHMM) \citep{Barajas:2017,Adam:2019}.
%which is comprised of two distinct, but related, HMMs that jointly model these time series. 
Next, we describe how to reduce fine-scale correlation by transforming raw observations into meaningful summary statistics. In particular, we apply a discrete Fourier transform (DFT) to the components of $Y^*$ using a moving window to summarize its periodic behaviour. Finally, we show how each of these models can be altered and combined to form a wide variety of new models.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Relaxing the conditional independence assumption with a CarHMM}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Transforming observations to account for fine-scale behaviour}
\label{subsec:STFT}

In many applications where data are collected at high frequencies, intricate dependency structures may arise within the fine-scale process which cannot be adequately modelled with any of the HMM-based models mentioned thus far. For example, the accelerometer data in Figure \ref{fig:data} sporadically displays clear periodic behaviour. To handle these additional fine-scale structures, we recommend replacing the fine-scale observations $Y_t^*$ with relevant statistics that summarize this non-Markovian behaviour. To maintain the temporal structure of the fine-scale process, summary statistics can be calculated from a moving window over $Y_t^*$. Subject-area experts are often required to determine the optimal window size and the specific summary statistics used. 
%
%This moving window approach complicates the indexing of $Y_t^*$. Suppose a window size of $h$ observations is selected. Moving forward, we re-index $Y^*$ such that the fine-scale process associated with curve $t$ is denoted as $Y^*_t$ as before. However, the $t^{*th}$ \textit{window} of $Y^*_t$ is now denoted as $Y^*_{t,t^*}$, and the $t^{**th}$ \textit{observation} of $Y^*_{t,t^*}$ is denoted as $Y^*_{t,t^*,t^{**}}$. As a result, there are a total of $T^*_t$ windows associated with $Y_t^*$, each of which are of length $h$. Note that this relabeling may significantly reduce the length of $Y^*_t$ depending upon the window size $h$ and the step-size between windows.
%

For our case study, we use the discrete Fourier transform (DFT) of a moving window of width $h$ across $Y^*_t$:
%
\begin{align*}
    DFT\{Y^*_{t,t^*}\}(k) = \sum_{n=0}^{h-1} Y^*_{t,t^*+n}\exp\left(-\frac{i 2\pi}{h} kn \right),
\end{align*}
%
for $k = 0, 1, \ldots, h-1$. If $Y^*_{t,t^*}$ is high-dimensional, then the DFT is taken component-wise. Next, we calculate $\Z_{t,t^*} \equiv \{\Zone_{t,t^*},\Ztwo_{t,t^*}\}$ for each moving window, where
%
\begin{equation}
    \label{eqn:z}
    \Zone_{t,t^*} \equiv \frac{1}{h}\sum_{n=0}^{h-1}Y^*_{t,t^*+n} \quad \text{and} \quad \Ztwo_{t,t^*} \equiv \sum_{k=1}^{\tilde{\omega}}\bigg|\bigg|DFT\{Y^*_{t,t^*}\}(k)\bigg|\bigg|^2.
\end{equation}
%
In words, $\Zone_{t,t^*}$ is the average value over the window and $\Ztwo_{t,t^*}$ is the squared 2-norm of the component of $\left(Y^*_{t,t^*}, \ldots, Y^*_{t,t^*+h-1}\right)$ that can be attributed to frequencies between $1$ and $\tilde{\omega}$ periods per window (by the Plancherel theorem). More intuitively, $\Ztwo_{t,t^*}$ corresponds to the ``wiggliness" of the acceleration data for dive $t$ within the window starting at time stamp $t^*$. The maximum frequency $\tilde{\omega}$ is a problem-specific tuning parameter which should be selected with the help of a subject-matter expert. Other DFT-based choices for window summary statistics are possible, including the dominant frequency and amplitude. To avoid introducing auto-correlation into the transformed observation sequence, we recommend down-sampling $\Z_t$ such that $\Z_t = \left(\Z_{t,1},\Z_{t,h+1},\Z_{t,2h+1},\ldots\right)$. The down-sampling avoids the artificial residual correlation that overlapping windows would introduce and reduces the dimension of the fine-scale process by a factor of $h$, allowing for faster model fitting. By summarizing the data using a relevant transformation, the method overall also reduces correlation in the raw observations due to the complex dependence structure. Figure \ref{fig:models}d represents the process of transforming $Y^*_{t}$ into $\Z_{t}$ within an HMM, including down-sampling.

@@

%so that the fine-scale emission distribution becomes $f^{(i)}\left(y_t,z^*_t;\theta^{(i)}\right)$, or more succinctly $f^{(i)}\left(y_t,z^*_t\right)$. The likelihood of the HMM-DFT is as follows:
%\begin{equation}
%    \calL_{\text{HMM-DFT}}(y,z^*;\theta,\Gamma) = \delta P(y_1,z^*_1;\theta) \prod_{t=2}^T \Gamma P(y_t,z^*_t;\theta) \mathbf{1}_N
%    \label{HMMDFT_likelihood}
%\end{equation}
%where $P(y_t,z^*_t;\theta)$ is an $N \times N$ diagonal matrix with $(i,i)^{th}$ entry equal to $f^{(i)}\left(y_t,z^*_t;\theta^{(i)}\right)$.

%It is possible to accommodate unequal time steps within $y_t^*$ by using the non-uniform discrete Fourier transform (NDFT). We do not describe the method in detail here, but the generalization is straightforward. See \citet{Bagchi:1999} for details.

@@

Possible candidates for the fine-scale model include any of the models described in the previous subsections (HMM, CarHMM, HHMM, and HHMM-DFT) in addition to many others not described here. 
For example, a continuous-time HMM (CTHMM) may be appropriate if observations are not equi-spaced in time \citep{Liu:2015}. In particular, \citet{Xu:2018} model biologging accelerometer data by incorporating a CTHMM into a hierarchical model similar to ours. However, they take a longitudinal approach whereas we use a coarse-scale HMM in our case study.
Further, \citet{Bebbington:2007} and \citet{Borchers:2013} model data sets with count onsets as observations, so they use variations on a Poisson process as their fine-scale model. If the fine-scale model is a simple Poisson process, this model reduces to a Markov-modulated Poisson process \citep{Fischer:1993}.
In addition, \citep{Langrock:2018} use B-splines to model several aspects of an HMM, and this modified model admits a penalized likelihood which can be incorporated into the hierarchical structure here. 
%For these and countless other possible fine-scale models, added flexibility comes at a cost of increased model complexity.

These examples are a few of many fine-scale models that can act as initial building blocks in a practitioner's toolbox to construct increasingly complex hierarchical models. As an example, consider a \textbf{CarHHMM}, in which the coarse-scale model is a traditional HMM and the fine-scale model is a CarHMM. Alternatively, one may construct an \textbf{HHHMM}, where three scales of observations are recorded ($Y$, $Y^*$, and $Y^{**}$) and traditional HMMs are used to model all three levels. It is clear that a a myriad of possible models can be built using this framework. However, these models quickly become complicated and should be constructed with care to achieve an adequate fit of the data while avoiding over-fitting and high computational costs.
%In the sections that follow, we construct several candidate models from the building blocks above to describe killer whale dive behaviour. We select one preferred model and compare it to several of the simpler candidate models to ensure its complexity is necessary.

>>>>

Hidden Markov models describe state-switching Markovian processes in discrete time and will be the core structure of the models we use for $Y$ and $Y^*$. We first detail the structure of a traditional HMM. We then review a variation described by \cite{Lawler:2019} called the conditionally auto-regressive hidden Markov model (CarHMM), which explicitly models auto-correlation between observations. Either the HMM or the CarHMM can be used to model a sequence of observations at one of the two scales: for the coarse scale sequence $Y$ or for each of the fine scale sequences, $Y_1^*,\ldots, Y_T^*$. To simultaneously model $Y$ and $Y^*$, we use the hierarchical hidden Markov model (HHMM) \citep{Barajas:2017,Adam:2019}.
%which is comprised of two distinct, but related, HMMs that jointly model these time series. 
Next, we describe how to reduce fine-scale correlation by transforming raw observations into meaningful summary statistics. In particular, we apply a discrete Fourier transform (DFT) to the components of $Y^*$ using a moving window to summarize its periodic behaviour. Finally, we show how each of these models can be altered and combined to form a wide variety of new models.

@@

\subsection{Relaxing the conditional independence assumption with a CarHMM}

@@

\subsection{Transforming observations to account for fine-scale behaviour}
\label{subsec:STFT}

In many applications where data are collected at high frequencies, intricate dependency structures may arise within the fine-scale process which cannot be adequately modelled with any of the HMM-based models mentioned thus far. For example, the accelerometer data in Figure \ref{fig:data} sporadically displays clear periodic behaviour. To handle these additional fine-scale structures, we recommend replacing the fine-scale observations $Y_t^*$ with relevant statistics that summarize this non-Markovian behaviour. To maintain the temporal structure of the fine-scale process, summary statistics can be calculated from a moving window over $Y_t^*$. Subject-area experts are often required to determine the optimal window size and the specific summary statistics used. For our case study, we use the discrete Fourier transform (DFT) of a moving window of width $h$ across $Y^*_t$:
%
\begin{align*}
    DFT\{Y^*_{t,t^*}\}(k) = \sum_{n=0}^{h-1} Y^*_{t,t^*+n}\exp\left(-\frac{i 2\pi}{h} kn \right),
\end{align*}
%
for $k = 0, 1, \ldots, h-1$. If $Y^*_{t,t^*}$ is high-dimensional, then the DFT is taken component-wise. Next, we calculate $\Z_{t,t^*} \equiv \{\Zone_{t,t^*},\Ztwo_{t,t^*}\}$ for each moving window, where
%
\begin{equation}
    \label{eqn:z}
    \Zone_{t,t^*} \equiv \frac{1}{h}\sum_{n=0}^{h-1}Y^*_{t,t^*+n} \quad \text{and} \quad \Ztwo_{t,t^*} \equiv \sum_{k=1}^{\tilde{\omega}}\bigg|\bigg|DFT\{Y^*_{t,t^*}\}(k)\bigg|\bigg|^2.
\end{equation}
%
In words, $\Zone_{t,t^*}$ is the average value over the window and $\Ztwo_{t,t^*}$ is the squared 2-norm of the component of $\left(Y^*_{t,t^*}, \ldots, Y^*_{t,t^*+h-1}\right)$ that can be attributed to frequencies between $1$ and $\tilde{\omega}$ periods per window (by the Plancherel theorem). More intuitively, $\Ztwo_{t,t^*}$ corresponds to the ``wiggliness" of the acceleration data for dive $t$ within the window starting at time stamp $t^*$. The maximum frequency $\tilde{\omega}$ is a problem-specific tuning parameter which should be selected with the help of a subject-matter expert. Other DFT-based choices for window summary statistics are possible, including the dominant frequency and amplitude. To avoid introducing auto-correlation into the transformed observation sequence, we recommend down-sampling $\Z_t$ such that $\Z_t = \left(\Z_{t,1},\Z_{t,h+1},\Z_{t,2h+1},\ldots\right)$. The down-sampling avoids the artificial residual correlation that overlapping windows would introduce and reduces the dimension of the fine-scale process by a factor of $h$, allowing for faster model fitting. By summarizing the data using a relevant transformation, the method overall also reduces correlation in the raw observations due to the complex dependence structure. Figure \ref{fig:models}d represents the process of transforming $Y^*_{t}$ into $\Z_{t}$ within an HMM, including down-sampling.

@@

@@

Possible candidates for the fine-scale model include any of the models described in the previous subsections (HMM, CarHMM, HHMM, and HHMM-DFT) in addition to many others not described here. 
For example, a continuous-time HMM (CTHMM) may be appropriate if observations are not equi-spaced in time \citep{Liu:2015}. In particular, \citet{Xu:2018} model biologging accelerometer data by incorporating a CTHMM into a hierarchical model similar to ours. However, they take a longitudinal approach whereas we use a coarse-scale HMM in our case study.
Further, \citet{Bebbington:2007} and \citet{Borchers:2013} model data sets with count onsets as observations, so they use variations on a Poisson process as their fine-scale model. If the fine-scale model is a simple Poisson process, this model reduces to a Markov-modulated Poisson process \citep{Fischer:1993}.
In addition, \citep{Langrock:2018} use B-splines to model several aspects of an HMM, and this modified model admits a penalized likelihood which can be incorporated into the hierarchical structure here.

These examples are a few of many fine-scale models that can act as initial building blocks in a practitioner's toolbox to construct increasingly complex hierarchical models. As an example, consider a \textbf{CarHHMM}, in which the coarse-scale model is a traditional HMM and the fine-scale model is a CarHMM. Alternatively, one may construct an \textbf{HHHMM}, where three scales of observations are recorded ($Y$, $Y^*$, and $Y^{**}$) and traditional HMMs are used to model all three levels. It is clear that a a myriad of possible models can be built using this framework. However, these models quickly become complicated and should be constructed with care to achieve an adequate fit of the data while avoiding over-fitting and high computational costs.

### 2020-09-01 Report/Sections/results.tex
Clean up report and delete old files

The data used in this study were collected on September 2, 2019 from 12:49 pm to 6:06 pm and consist of depth and acceleration in three orthogonal directions. Observations were collected at a rate of 50 Hz using
% an accelerometer and time-depth recorder.
[NAME OF TAGS - do you know Sarah?]. The acceleration was measured in three dimensions, which together represent the complete range of movement of an animal (forward/backward, upward/downward, right/left). Tri-axial acceleration readings are common in these types of tags \citep{Cade:2017,Fehlmann:2017,Wright:2017}. Tagging the killer whale caused anomalous behaviour before 1:20 pm and after 6:00 pm, so observations in these time periods are ignored. We also ignore all dives between 2:25 pm and 2:37 pm and between 4:07 pm and 5:07 pm, when the tagging technology malfunctioned. We smooth the accelerometer and depth data by taking a moving average within a window of $1/10^{th}$ of a second. We then define a killer whale ``dive" as any continuous section of data that occurs below 0.5 meters in depth and lasts for at least 10 seconds. Data are preprocessed in part with the \textit{divebomb} package in Python \citep{Nunes:2018}. The preprocessed data contain a total of 267 dives, all of which are displayed in Figure \ref{fig:data}. Each dive is treated as one curve, and the sequence of dives makes up the coarse-scale process. Specifically, the vector of observed coarse-scale observations $y = \Big\{y_1,\ldots,y_{267}\Big\}$ is a time series of dive durations in seconds ($s$) with corresponding vector of unobserved dive types $x = \Big\{x_1,\ldots,x_{267}\Big\}$. For dive $t$, the fine-scale observations are contained in $y^*_{t}$, which is a vector of the within-dive acceleration data in meters per second squared ($m/s^2$). The corresponding vector of subdive states is $x^*_{t}$. Naturally, $y^* = \Big\{y^*_1,\ldots,y^*_{267}\Big\}$ and $x^* = \Big\{x^*_1,\ldots,x^*_{267}\Big\}$ is the corresponding set of subdive states.

@@

To choose how to model the fine-scale observations (acceleration), we again begin with determining model structure. In contrast to the coarse-scale, the fine-scale acceleration data exhibit significant sinusoidal behaviour. Thus, we transform each fine-scale observation sequence $y_t^*$ into $z_t^*$ using Equation (\ref{eqn:z}). We select a window size of two seconds since it appears be the smallest window size that comfortably captures at least one full period of each ``wiggle" in the acceleration data.
%
We calculate $\Ztwo_{t,t^*}$ by summing the first $\tilde \omega = 10$ Fourier modes of each window, which corresponds to maximum frequencies of 5 Hz for each axis of acceleration. We also down-sample each observation $z^*_t$ such that $z^*_t = \Big\{z^*_{t,1},z^*_{t,101},z^*_{t,201},\ldots\Big\}$ to avoid overlapping windows. The resulting sequence of observations is $z^* = \Big\{z^*_1,\ldots,z^*_{267}\Big\}$, where $z^*_t = \Big\{z^*_{t,1},z^*_{t,101},\ldots\Big\}$ and $z^*_{t,t^*} = \{a^*_{t,t^*},w^*_{t,t^*}\}$. We have that $a^*_{t,t^*}$ is a three dimensional vector of component-wise average acceleration and $w^*_{t,t^*}$ is a scalar describing the ``wiggliness" of a given window.
%
After transforming the raw acceleration data, there is still strong auto-correlation within each axis of $\Zone_{t,t^*}$ prior to fitting the model (see section 2.1 of the supplementary material). Therefore, we assume that all three components of $\Zone_{t,t^*}$, $t^* = 1,101,\ldots$ follow the CarHMM as defined in section 2.2. 
%
%$$\mathbb{E}(\Zone_{t,t^*}|\\Zone_{t,t^*-1} = \mathbf{z}, X^*_{t,t^*} = i) = \phi_A^{*(i)} \mathbf{z} + (1-\phi_A^{*(i)}) \mathbf{\mu}_A^{*(i)},$$
%$$\mathbb{V}(\Zone_{t,t^*}|\Zone_{t,t^*-1} = z,X^*_{t,t^*} = i) = \text{diag}\left[\left(\mathbf{\sigma}_A^{*(i)}\right)^2\right].$$
%
While the $\Ztwo_{t,t^*}$ observations also exhibit some auto-correlation prior to fitting the model, it is less strong than that of $\Zone_{t,t^*}$, and 
%the biological interpretation of auto-correlation between observations of $\Ztwo_{t,t^*}$ is less clear. In addition, 
much of the auto-correlation evident from the lag plot may be explained by subsequent observations having the same subdive state. Therefore, a simple HMM structure is used to model $\Ztwo_{t,t^*}$ for all $t$ and $t^*$. 
%
%$$\mathbb{E}(\Ztwo_{t,t^*}|\Zone_{t,t^*-1} = z,X^*_{t,t^*} = i) = \mu_W^{*(i)}$$
%$$\mathbb{V}(\Ztwo_{t,t^*}|\Zone_{t,t^*-1} = z,X^*_{t,t^*} = i) = \left(\sigma_W^{*(i)}\right)^2.$$
%

@@

The estimated probability transition matrices and associated stationary distributions are
%
$$\hat \Gamma = \begin{pmatrix} 
0.788 & 0.212 \\
0.809 & 0.191
\end{pmatrix} \text{ and }$$
$$\hat \delta = \begin{pmatrix} 0.792 & 0.208 \end{pmatrix}$$
%
for the transitions between dives. Within dives, we have
$$\hat \Gamma^{*(1)} = \begin{pmatrix} 
0.679 & 0.321 & 0.000 \\
0.038 & 0.904 & 0.058 \\
0.000 & 0.232 & 0.768
\end{pmatrix}, \qquad 
\hat \Gamma^{*(2)} = \begin{pmatrix} 
0.859 & 0.141 & 0.000 \\
0.114 & 0.841 & 0.045 \\
0.000 & 0.216 & 0.784
\end{pmatrix},$$
$$\hat \delta^{*(1)} = \begin{pmatrix} 0.087 & 0.731 & 0.182 \end{pmatrix}, \enspace \text{and} \enspace \hat \delta^{*(2)} = \begin{pmatrix} 0.401 & 0.496 & 0.103 \end{pmatrix}$$
%
for dive types 1 and 2.
The estimated coarse-scale stationary distribution, $\hat{\delta}$, indicates that about 79\% of dives are short dives of type 1. The estimated coarse-scale transition probability matrix $\hat \Gamma$ further indicates that the whale performs an average of 4.72 short type 1 dives before switching to dive type 2 and an average of 1.24 longer type 2 dives before switching back to dive type 1. This finding is consistent with those of \citet{Tennessen:2019b} and \citet{Williams:2009}, both of whom describe common bouts of short resting dives before a killer whale performs a longer, more energy-intensive deep dive. The fine-scale stationary distributions $\hat{\delta}^{*(i^*)}$ indicate that this killer whale is in the less active subdive state 1 much more often when performing long deep dives compared to short shallow dives (40\% of the time for dive type 1 and 9\% of the time for dive type 2). Less active swimming behaviour is consistent with the need for marine mammals to conserve energy when diving at depth and holding breath for long periods of time \citep{Williams:1999,Hastie:2006}. Figure \ref{fig:labeled_dives} shows the decoded dive behaviour of six selected dives. Section 2.2 of the supplementary material also shows the estimated probability of each dive and subdive state given the data.

@@

We use two visual tools to evaluate the CarHHMM-DFT: pseudo-residual plots and empirical histograms. 
%A pseudo-residual is the marginal cumulative distribution function of an observation conditioned on all other observations \citep{Zucchini:2016}. To easily visualize outliers, pseudo-residuals are often passed through the quantile function of the standard Normal distribution.
Mathematically, the pseudo-residual of an observation $y_t$ for a traditional HMM is equal to $\Phi^{-1} \left(Pr(Y_t < y_t|\{Y_1,\ldots,Y_T\}/\{Y_t\}) \right)$, where $\Phi$ is the cumulative distribution function of a standard Normal distribution. If the model is correct, then all pseudo-residuals are independent and follow a standard Normal distribution. Histograms of the pseudoresiduals mostly support that this model is well-specified. One exception is $\Ztwo_{t,t^*}$, whose pseudo-residuals are noticeably right-skewed (Figure \ref{fig:pseudoresids}). This implies that the true distribution of $\Ztwo_{t,t^*}$ may follow a heavier-tailed distribution than the gamma distribution used in the case study. See section 2 of the supplementary material for pseudo-residual plots for all observations.

We also plot histograms of dive duration corresponding to each dive type in Figure \ref{fig:empirical_dist}. Each observed dive duration is weighted by the decoded probability that it corresponds to a particular dive type. This results in two histograms - one corresponding to dive type 1 and another corresponding to dive type 2. Each histograms is then plotted with the corresponding emission distribution learned by the CarHHMM-DFT (see Figure \ref{fig:empirical_dist}). See section 2 of the supplementary material for analogous histograms corresponding to the fine-scale observations $z^*_{t,t^*}$. Our results again mostly show that the CarHHMM-DFT fits the data well, but there are some exceptions. In particular, $\Ztwo_{t,t^*}$ is right-skewed and $\Zone_{t,t^*}$ has heavy tails corresponding to subdive state 3, indicating the existence of rare events corresponding to exceptionally violent thrashing of the killer whale. These outliers are potential subjects for future study and may indicate biologically relevant phenomena such as prey capture \citep{Tennessen:2019a}.

@@

Finally, the CarHHMM does not use the ``wiggliness" of the acceleration data as an observation, so it regularly fails to pick up obvious behavioural changes corresponding to the periodicity shown in Figure \ref{fig:labeled_dives}. This fact alone essentially disqualifies the CarHHMM as a viable model for these data. The pseudoresiduals corresponding to the CarHHMM also appear to be less heavy-tailed than a normal distribution. See section 2 of the supplementary material for a more complete set of results for each of the candidate models.

>>>>

The data used in this study were collected on September 2, 2019 from 12:49 pm to 6:06 pm and consist of depth and acceleration in three orthogonal directions. Observations were collected at a rate of 50 Hz using Customizable Animal Solutions Tracking (CATs) tags (model number ###). The acceleration was measured in three dimensions, which together represent the complete range of movement of an animal (forward/backward, upward/downward, right/left). Tri-axial acceleration readings are common in these types of tags \citep{Cade:2017,Fehlmann:2017,Wright:2017}. Tagging the killer whale caused anomalous behaviour before 1:20 pm and after 6:00 pm, so observations in these time periods are ignored. We also ignore all dives between 2:25 pm and 2:37 pm and between 4:07 pm and 5:07 pm, when the tagging technology malfunctioned. We smooth the accelerometer and depth data by taking a moving average within a window of $1/10^{th}$ of a second. We then define a killer whale ``dive" as any continuous section of data that occurs below 0.5 meters in depth and lasts for at least 10 seconds. Data are preprocessed in part with the \textit{divebomb} package in Python \citep{Nunes:2018}. The preprocessed data contain a total of 267 dives, all of which are displayed in Figure \ref{fig:data}. Each dive is treated as one curve, and the sequence of dives makes up the coarse-scale process. Specifically, the vector of observed coarse-scale observations $y = \Big\{y_1,\ldots,y_{267}\Big\}$ is a time series of dive durations in seconds ($s$) with corresponding vector of unobserved dive types $x = \Big\{x_1,\ldots,x_{267}\Big\}$. For dive $t$, the fine-scale observations are contained in $y^*_{t}$, which is a vector of the within-dive acceleration data in meters per second squared ($m/s^2$). The corresponding vector of subdive states is $x^*_{t}$. Naturally, $y^* = \Big\{y^*_1,\ldots,y^*_{267}\Big\}$ and $x^* = \Big\{x^*_1,\ldots,x^*_{267}\Big\}$ is the corresponding set of subdive states.

@@

To choose how to model the fine-scale observations (acceleration), we again begin with determining model structure. In contrast to the coarse-scale, the fine-scale acceleration data exhibit significant sinusoidal behaviour. Thus, we transform each fine-scale observation sequence $y_t^*$ into $z_t^*$ using Equation (\ref{eqn:z}). We select a window size of two seconds since it appears be the smallest window size that comfortably captures at least one full period of each ``wiggle" in the acceleration data.
%
We calculate $\Ztwo_{t,t^*}$ by summing the first $\tilde \omega = 10$ Fourier modes of each window, which corresponds to maximum frequencies of 5 Hz for each axis of acceleration. We also down-sample each observation $z^*_t$ such that $z^*_t = \Big\{z^*_{t,1},z^*_{t,101},z^*_{t,201},\ldots\Big\}$ to avoid overlapping windows. The resulting sequence of observations is $z^* = \Big\{z^*_1,\ldots,z^*_{267}\Big\}$, where $z^*_t = \Big\{z^*_{t,1},z^*_{t,101},\ldots\Big\}$ and $z^*_{t,t^*} = \{a^*_{t,t^*},w^*_{t,t^*}\}$. We have that $a^*_{t,t^*}$ is a three dimensional vector of component-wise average acceleration and $w^*_{t,t^*}$ is a scalar describing the ``wiggliness" of a given window.
%
After transforming the raw acceleration data, there is still strong auto-correlation within each axis of $\Zone_{t,t^*}$ prior to fitting the model (see section 1.1 of the supplementary material). Therefore, we assume that all three components of $\Zone_{t,t^*}$, $t^* = 1,101,\ldots$ follow the CarHMM as defined in section 2.2. 
%
%$$\mathbb{E}(\Zone_{t,t^*}|\\Zone_{t,t^*-1} = \mathbf{z}, X^*_{t,t^*} = i) = \phi_A^{*(i)} \mathbf{z} + (1-\phi_A^{*(i)}) \mathbf{\mu}_A^{*(i)},$$
%$$\mathbb{V}(\Zone_{t,t^*}|\Zone_{t,t^*-1} = z,X^*_{t,t^*} = i) = \text{diag}\left[\left(\mathbf{\sigma}_A^{*(i)}\right)^2\right].$$
%
While the $\Ztwo_{t,t^*}$ observations also exhibit some auto-correlation prior to fitting the model, it is less strong than that of $\Zone_{t,t^*}$, and 
%the biological interpretation of auto-correlation between observations of $\Ztwo_{t,t^*}$ is less clear. In addition, 
much of the auto-correlation evident from the lag plot may be explained by subsequent observations having the same subdive state. Therefore, a simple HMM structure is used to model $\Ztwo_{t,t^*}$ for all $t$ and $t^*$. 
%
%$$\mathbb{E}(\Ztwo_{t,t^*}|\Zone_{t,t^*-1} = z,X^*_{t,t^*} = i) = \mu_W^{*(i)}$$
%$$\mathbb{V}(\Ztwo_{t,t^*}|\Zone_{t,t^*-1} = z,X^*_{t,t^*} = i) = \left(\sigma_W^{*(i)}\right)^2.$$
%

@@

The estimated probability transition matrices and associated stationary distributions are
%
$$\hat \Gamma = \begin{pmatrix} 
0.788 & 0.212 \\
0.809 & 0.191
\end{pmatrix} \text{ and }$$
$$\hat \delta = \begin{pmatrix} 0.792 & 0.208 \end{pmatrix}$$
%
for the transitions between dives. Within dives, we have
$$\hat \Gamma^{*(1)} = \begin{pmatrix} 
0.679 & 0.321 & 0.000 \\
0.038 & 0.904 & 0.058 \\
0.000 & 0.232 & 0.768
\end{pmatrix}, \qquad 
\hat \Gamma^{*(2)} = \begin{pmatrix} 
0.859 & 0.141 & 0.000 \\
0.114 & 0.841 & 0.045 \\
0.000 & 0.216 & 0.784
\end{pmatrix},$$
$$\hat \delta^{*(1)} = \begin{pmatrix} 0.087 & 0.731 & 0.182 \end{pmatrix}, \enspace \text{and} \enspace \hat \delta^{*(2)} = \begin{pmatrix} 0.401 & 0.496 & 0.103 \end{pmatrix}$$
%
for dive types 1 and 2.
The estimated coarse-scale stationary distribution, $\hat{\delta}$, indicates that about 79\% of dives are short dives of type 1. The estimated coarse-scale transition probability matrix $\hat \Gamma$ further indicates that the whale performs an average of 4.72 short type 1 dives before switching to dive type 2 and an average of 1.24 longer type 2 dives before switching back to dive type 1. This finding is consistent with those of \citet{Tennessen:2019b} and \citet{Williams:2009}, both of whom describe common bouts of short resting dives before a killer whale performs a longer, more energy-intensive deep dive. The fine-scale stationary distributions $\hat{\delta}^{*(i^*)}$ indicate that this killer whale is in the less active subdive state 1 much more often when performing long deep dives compared to short shallow dives (40\% of the time for dive type 1 and 9\% of the time for dive type 2). Less active swimming behaviour is consistent with the need for marine mammals to conserve energy when diving at depth and holding breath for long periods of time \citep{Williams:1999,Hastie:2006}. Figure \ref{fig:labeled_dives} shows the decoded dive behaviour of six selected dives. Section 1.3 of the supplementary material also shows the probability of each dive type and subdive state given the data and the fitted model.

@@

We use two visual tools to evaluate the CarHHMM-DFT: pseudoresidual plots and empirical histograms. Mathematically, the pseudoresidual of an observation $y_t$ for a traditional HMM is equal to $\Phi^{-1} \left(Pr(Y_t < y_t|\{Y_1,\ldots,Y_T\}/\{Y_t\}) \right)$, where $\Phi$ is the cumulative distribution function of a standard Normal distribution. If the model is correct, then all pseudoresiduals are independent and follow a standard Normal distribution. Histograms of the pseudoresiduals mostly support that this model is well-specified. One exception is $\Ztwo_{t,t^*}$, whose pseudoresiduals are noticeably right-skewed (Figure \ref{fig:pseudoresids}). This implies that the true distribution of $\Ztwo_{t,t^*}$ may follow a heavier-tailed distribution than the gamma distribution used in the case study. See sections 1.4 - 1.6 of the supplementary material for pseudoresidual plots for all observations.

We also plot histograms of dive duration corresponding to each dive type in Figure \ref{fig:empirical_dist}. Each observed dive duration is weighted by the decoded probability that it corresponds to a particular dive type. This results in two histograms - one corresponding to dive type 1 and another corresponding to dive type 2. Each histograms is then plotted with the corresponding emission distribution learned by the CarHHMM-DFT (see Figure \ref{fig:empirical_dist}). See sections 1.4 - 1.6 of the supplementary material for analogous histograms corresponding to the fine-scale observations $z^*_{t,t^*}$. Our results again mostly show that the CarHHMM-DFT fits the data well, but there are some exceptions. In particular, $\Ztwo_{t,t^*}$ is right-skewed and $\Zone_{t,t^*}$ has heavy tails corresponding to subdive state 3, indicating the existence of rare events corresponding to exceptionally violent thrashing of the killer whale. These outliers are potential subjects for future study and may indicate biologically relevant phenomena such as prey capture \citep{Tennessen:2019a}.

@@

Finally, the CarHHMM does not use the ``wiggliness" of the acceleration data as an observation, so it regularly fails to pick up obvious behavioural changes corresponding to the periodicity shown in Figure \ref{fig:labeled_dives}. This fact alone essentially disqualifies the CarHHMM as a viable model for these data. The pseudoresiduals corresponding to the CarHHMM also appear to be less heavy-tailed than a normal distribution. See section 1 of the supplementary material for a more complete set of results for each of the candidate models.
